### 2022-09-14 main/chapter_searches.tex
add some awkward words and drafts on cls; update figures with sky blue to be clear in greyscale

Statistical data analysis suffers from dogma --- prescriptions and recipes that
students are taught to memorize and follow without knowing their reasons or
limitations.
From a standard undergraduate physics education, my experience has been
error propagation formulae,
least-squares fitting routines,
chi-squared per degree of freedom,
and myriad statistical tests and confidence intervals.
Many of these are effective approximations, but all also fail if misused
outside their domain.%
\footnote{%
Multiplicative error propagation is an example.
For $a = \mu_a \pm \sigma_a$ and $b = \mu_b \pm \sigma_b$,
the standard perturbative approximation is
\(
\widetilde{\sigma}_{ab}
= \mu_a \mu_b \sqrt{\sigma_a^2/\mu_a^2 + \sigma_b^2/\mu_b^2}
% = \sigma_a^2 \mu_b^2 + \sigma_b^2 \mu_a^2
\),
but a short calculation with their moments derives the exact result
\(
\sigma_{ab}
= \sqrt{(\mu_a^2 + \sigma_a^2)(\mu_b^2 + \sigma_b^2) - \mu_a^2\mu_b^2}
% = \sigma_a^2 \mu_b^2 + \sigma_b^2 \mu_a^2 + \sigma_a^2 \sigma_b^2
\)
if $a$ and $b$ are independent.
Since $\sigma_{ab}^2 = \widetilde{\sigma}_{ab}^2 + \sigma_a^2 \sigma_b^2$, this
propagation formula works only for small relative errors.
}
All effective methods of data analysis can be derived upwards from the roots of
probability and decision theory, and as elsewhere in physics, by knowing the
derivations we can also know when each approximation should work and when it
needs replacing.

@@

Logic is an important application of probability for which $a$ and $b$ are
interpreted as logical propositions, and
$\prob{b}{a}$ is the degree to which $a$ implies $b$, with the limiting cases
that $\prob{b}{a} = 1$ when $a$ implies $b$, and $\prob{b}{a} = 0$ when $a$
contradicts $b$.
Indeed, Cox's theorem derives the laws of probability by requiring compatibility
with Boolean logic~\cite{
cox1946probability,
cox1961algebra,
garrett1998nand,
jaynes2003probability,
keynes1920treatise
}.
Probability works for logic, and it works \emph{for tea, too}.

@@

\subsection{Probability, applied}
\begin{figure}[tp]
\centering
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/searches_abacus_likelihood.pdf}
\caption{%
Likelihood: $\mathrm{Poisson}(0\mid x)$ in
(left) $x$ and
(right) $u = \log x$%
}
\end{subfigure}
\\[.5em]
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/searches_abacus_prior.pdf}
\caption{%
Prior density: exponential in
(left) $x$ and
(right) $u = \log x$%
}
\end{subfigure}
\caption[
Likelihoods and prior densities' different transformations
]{%
Likelihoods and prior densities' different transformations with coordinates.
The $\mathrm{Poisson}(0\mid x)$ likelihood and exponential $e^{-x}\mathrm{d}x$
prior have the same form, but the prior density changes to preserve its unit
area when expressed relative to $u = \log x$.
\\[.5em]
Relative to $x$, both are identical (left).
On changing to $\log$ coordinates (right), the likelihood slides horizontally,
but the prior density changes value to preserve area.
Both maximum likelihood points are equivalent (maximum likelihoods are robust),
but the maximum prior density moves (maximum densities are slippery).
\\[.5em]
Particle lifetimes follow exponential distributions, and their maximum
densities are at zero time.
It is misleading, however, to describe their most probable lifetimes as zero
---
the `most probable' $\log$ lifetime is not $\log 0$, but equals the mean
lifetime.
\\[.5em]
For zero Poisson data, the maximum likelihood rate is zero, but zero might not
be the most likely rate (if, for example, an experiment has known backgrounds).
}
\label{fig:searches_abacus}
\end{figure}
% science snake
Science involves making predictions, testing those predictions against data,
and reporting what is learned from the results.
Various probabilities are involved in this prediction and learning that have
conventional names to communicate what we intend to do with them.
Most important for prediction is a \textbf{likelihood}
\begin{align}
L(a) &= \prob{b}{a} \\
\label{eqn:searches_likelihood}
\intertext{%
that explores the probabilities assigned to a fixed result $b$ from different
contexts $a$, or how well different hypotheses predicted a result.
A likelihood is clearly a probability, but it is not a probability distribution
--- only the left hand argument $b$ forms a normalized distribution.
Alternatively, fixing the context gives a \textbf{prior}%
}
\pi(b) &= \prob{b}{a}
\label{eqn:searches_prior}
\end{align}
for fixed $a$, which is a normalized distribution over disjoint states $b$.

@@

Neyman and Pearson were also involved in the development of
`confidence intervals', which are essentially ranges of parameters that are
accepted by the \pvalue\ decision rules~\cite{
clopper1934confidence,
neyman1935Intervals,
Neyman1937Outline
}.%
\footnote{%
Confidence are usually introduced by endpoints containing constrained
probability mass, but the arbitrary choice of test statistic can always reduce
that to this \pvalue\ form.%
}
A parameter corresponds to an hypothesis that assigns a prior distribution to
plausible data, and those data for which an obeisant researcher accepts the
hypothesis have probabilities that sum to $\gamma = 1 - \alpha$.
This $\gamma$ is known as the `confidence level'.
Such a rule selects a set of favoured data for each hypothesis, and the
confidence interval (when it can be constructed) contains hypotheses for which
the observed data are in that favoured set.

@@

\TODO{positive message:
not meant only to disparage our lies, but to indicate that these explanations
arise to fill a hole of confusion that we can fill by reporting interpretable
statistics such as $\log$ likelihood ratios.
}

\begin{figure}[tp]
\centering
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/searches_cls_plots_with_pvals_n.pdf}
\caption{%
(left) \pvalues\ for Poisson data,
and (right) $\log$ likelihood ratios and $\log$ $\cls$ for
background and signal-plus-background models.%
}
\end{subfigure}
\caption[
Analysis of a minimal Poisson modelling example in the natural order.
test statistics
]{%
Analysis of a minimal Poisson modelling example with $t(n) = \pm n$
test statistics; $b = 4.3$ and $s = 3.0$.
The $n$ test statistic is Neyman-Pearson optimal because it is monotonically
(in fact linearly) related to the likelihood ratio.
Each $\cls$ is a ratio between signal-plus-background and background-only
\pvalues\ as defined in Equation~\ref{eqn:searches_cls}.
By reducing the observed data to an inequality, the $p$-value selectively
censors some of its information.
}
\label{fig:searches_sb_n}
\end{figure}

\begin{figure}[tp]
\centering
\begin{subfigure}{0.5\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/searches_cls_plots_with_pvals_order.pdf}
\caption{%
Ordering the data by a maximum likelihood ratio test statistic.
}
\end{subfigure}
\\[.5em]
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/searches_cls_plots_with_pvals_t.pdf}
\caption{%
(left) \pvalues\ for Poisson data,
and (right) $\log$ likelihood ratios and $\log$ $\cls$ for
background and signal-plus-background models.%
}
\end{subfigure}
\caption[
Analysis of a minimal Poisson modelling example with maximum likelihood ratio
test statistics
]{%
Analysis of a minimal Poisson modelling example with maximum likelihood ratio
test statistics $t(n) = \pm \Lambda(n)$,
where $\Lambda(n) = L(b)/\mathrm{max}_x\,L(x)$; $b = 4.3$ and $s = 3.0$.
This test statistic jumbles up the data into an that makes $n=11$ appear to be
weaker support for the signal than $n=9$.
Each $\cls$ is a ratio between signal-plus-background and background-only
\pvalues\ as defined in Equation~\ref{eqn:searches_cls}.
By reducing the observed data to an inequality, the $p$-value selectively
censors some of its information.
}
\label{fig:searches_sb_t}
\end{figure}

Figure~\ref{fig:searches_sb_n}
Figure~\ref{fig:searches_sb_t}

\TODO{n test stat and history first?}
For example, maximum likelihood ratios (and variations on their theme) are
useful and commonly applied, but they can give \pvalues\ that appear
pathological --- consider a Poisson background ($b$) added to an
unconstrained signal yield ($s$).
A sensible interpretation would find that `excess' data well above the
background expectation favour the background-plus-signal hypothesis ($s + b$)
---
but if the signal is allowed to be negative, then a `deficit' of data below the
background can have a maximum likelihood ratio that favours $s + b$.
This scrambles excesses and deficits together in the same tail, so both get
small \pvalues, and so a small \pvalue\ implies neither a preference for the
signal hypothesis nor incompatibility with the background.

As discussed in Section~\ref{sec:searches_fisher}, \pvalues\ are likelihoods
on obfuscated data, so they are interpretable in likelihood ratios, and such
\pvalue\ ratios are in fact standard practice at the LHC when setting upper
limits on disfavoured signal models.
The standard ratio is named $\cls$~\cite{
read2000modified,
Read2002cls,
Murray2010heretic,
pdg2022ynf
}
and defined as
\begin{equation}
\label{eqn:searches_cls}
\cls
=
\frac{P(s + b)}{P(b)\hphantom{s +\;}}
=
\frac{
\prob{t \geq \tilde{t}}{s + b}
}{
\prob{t \geq \tilde{t}}{b}\hphantom{s +\;}
}
,
\end{equation}
in which the individual \pvalues\ are labelled
$\clb = P(b)$ (for background only) and
$\clsb = P(s + b)$ (for signal plus background).%
\footnote{%
Despite the $\mathrm{CL}$ labels, these ($\cls$, $\clsb$, and $\clb$) are
not confidence levels ---
a confidence level $\gamma$ is a decision boundary used, for example,
to assign limits by solving for $x$ such that
$\gamma = 1 - P(x)$~\cite{pdg1996}.
In particle physics, \pvalues\ have sometimes been labelled $\mathrm{CL}$ since
at least 1984~\cite{pdg1984}, although textbooks from the '70s (which are cited
by~\cite{pdg1984}) do make the distinction~\cite{
eadie1971statistical,
frodesen1979probability
}.%
}
Some literature also defines $\clb = 1 - p_b$~\cite{
Cowan:2010js,
pdg2022ynf
},
but $p_b$ may not quite be a \pvalue\ if it excludes equality with the observed
test statistic.

% history: bayes uniform prior in special cases
% just an approximation
% use for Higgs at LEP to avoid spurious rejection of bkg
LEP Higgs boson searches~\cite{
read1997optimal,
bock1998lower,
etde1998prospects,
junk1999confidence,
lep2000searches
}

credible intervals in simple Poisson or Gaussian cases~\cite{
Helene1983upper,
pdg1988,
pdg2022ynf
}

origins~\cite{
pdg1988,
zech1988cls,
bock1998lower,
read1997optimal,
Read2002cls,
Murray2010heretic
}

\TODO{CL \pvalue~\cite{pdg1984, murray1997use}}
\TODO{Helene~\cite{pdg1988}}
\TODO{LEP~\cite{etde1998prospects, lep2000searches, lep2003search}}

\TODO{Alas, slightly different test statistics }

>>>>

Statistical data analysis suffers from dogma --- prescriptions and recipes that
students are taught to memorize and follow without knowing their reasons or
limitations.
From a standard undergraduate physics education, my experience has been
error propagation formulae,
least-squares fitting routines,
chi-squared per degree of freedom,
and myriad statistical tests and confidence intervals.
Many of these are effective approximations, but all also fail if misused
outside their domain.%
\footnote{%
Multiplicative error propagation is an example.
For $a = \mu_a \pm \sigma_a$ and $b = \mu_b \pm \sigma_b$,
the standard perturbative approximation is
\(
\widetilde{\sigma}_{ab}
= \mu_a \mu_b \sqrt{\sigma_a^2/\mu_a^2 + \sigma_b^2/\mu_b^2}
% = \sigma_a^2 \mu_b^2 + \sigma_b^2 \mu_a^2
\),
but a calculation with their moments derives the exact result
\(
\sigma_{ab}
= \sqrt{(\mu_a^2 + \sigma_a^2)(\mu_b^2 + \sigma_b^2) - \mu_a^2\mu_b^2}
% = \sigma_a^2 \mu_b^2 + \sigma_b^2 \mu_a^2 + \sigma_a^2 \sigma_b^2
\)
if $a$ and $b$ are independent.
Since $\sigma_{ab}^2 = \widetilde{\sigma}_{ab}^2 + \sigma_a^2 \sigma_b^2$, the
propagation formula works only for small relative errors.
}
All effective methods of data analysis can be derived upwards from the roots of
probability and decision theory, and as elsewhere in physics, by knowing the
derivations we can also know when each approximation should work and when it
needs replacing.

@@

Logic is an important application of probability, for which $a$ and $b$ are
interpreted as logical propositions and
$\prob{b}{a}$ is the degree to which $a$ implies $b$, with the limiting cases
that $\prob{b}{a} = 1$ when $a$ implies $b$, and $\prob{b}{a} = 0$ when $a$
contradicts $b$.
Indeed, Cox's theorem derives the laws of probability by requiring compatibility
with Boolean logic~\cite{
cox1946probability,
cox1961algebra,
garrett1998nand,
jaynes2003probability,
keynes1920treatise
}.
Probability works for logic, and it works \emph{for tea, too}.

Probability \emph{distributions} $\prob{b}{a}$ spread unit probability mass
across states $b$, and need not be random distributions of fluctuating
$b$s~\cite{jaynes2003probability,frankfurt2005on}, although random processes
are useful applications.

@@

\subsection{Probability, applied}
\begin{figure}[tp]
\centering
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/searches_abacus_likelihood.pdf}
\caption{%
Likelihood: $\mathrm{Poisson}(0\mid x)$ in
(left) $x$ and
(right) $u = \log x$%
}
\end{subfigure}
\\[.5em]
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/searches_abacus_prior.pdf}
\caption{%
Prior density: exponential in
(left) $x$ and
(right) $u = \log x$%
}
\end{subfigure}
\caption[
Likelihoods and prior densities' different transformations
]{%
Likelihoods and prior densities' different transformations with coordinates.
The $\mathrm{Poisson}(0\mid x)$ likelihood and exponential $e^{-x}\mathrm{d}x$
prior have the same form, but the prior density changes to preserve its unit
area when expressed relative to $u = \log x$.
\\[.5em]
Relative to $x$, both are identical (left).
On changing to $\log$ coordinates (right), the likelihood slides horizontally,
but the prior density changes value to preserve area.
Both maximum likelihood points are equivalent (maximum likelihoods are robust),
but the maximum prior density moves (maximum densities are slippery).
\\[.5em]
Particle lifetimes follow exponential distributions, and their maximum
densities are at zero time.
It is misleading, however, to describe their most probable lifetimes as zero
---
the `most probable' $\log$ lifetime is not $\log 0$, but equals the mean
lifetime.
\\[.5em]
For zero Poisson data, the maximum likelihood rate is zero, but zero might not
be the most likely rate (if, for example, an experiment has known backgrounds).
}
\label{fig:searches_abacus}
\end{figure}
% science snake
Science involves making predictions, testing those predictions against data,
and reporting what is learned from the results.
Various probabilities are involved in this prediction and learning that have
conventional names to communicate what we intend to do with them.
Most important for prediction is a \textbf{likelihood}
\begin{align}
L(a) &= \prob{b}{a} \\
\label{eqn:searches_likelihood}
\intertext{%
that explores the probabilities assigned to a fixed result $b$ from different
contexts $a$, or how well different hypotheses predicted a result.
A likelihood is clearly a probability, but it is not a probability distribution
--- only the left hand argument $b$ forms a normalized distribution.
Alternatively, fixing the context gives a \textbf{prior}%
}
\pi(b) &= \prob{b}{a}
\label{eqn:searches_prior}
\end{align}
for fixed $a$, which is a normalized probability distribution over $b$.

@@

Neyman and Pearson were also involved in the development of
`confidence intervals', which are ranges of parameters that are
accepted by the \pvalue\ decision rules~\cite{
clopper1934confidence,
neyman1935Intervals,
Neyman1937Outline
}.%
\footnote{%
Confidence are usually introduced by endpoints containing constrained
probability mass, but this \pvalue\ form can always satisfy that through the
free choice of test statistic and extends naturally beyond one dimension.%
}
A parameter corresponds to an hypothesis that assigns a prior distribution to
plausible data, and those data for which an obeisant researcher accepts the
hypothesis have probabilities that sum to $\gamma = 1 - \alpha$.
This $\gamma$ is known as the `confidence level'.
Such a rule selects a set of favoured data for each hypothesis, and the
confidence interval (when it can be constructed) contains hypotheses for which
the observed data are in that favoured set.

@@

Such misrepresentations of our results reinforces the need for clearer
replacements, which $\log$ likelihood ratios can provide.
And standard practice at the LHC contains the beginnings of such
replacements under the name $\cls$~\cite{
read2000modified,
Read2002cls,
junk1999confidence,
read1997optimal,
bock1998lower,
etde1998prospects,
lep2000searches,
lep2003search,
Murray2010heretic,
CMS-NOTE-2011-005,
pdg2022ynf
},
which is used for setting upper limits on disfavoured signal yields $s$
above background yields $b$;
\begin{equation}
\label{eqn:searches_cls}
\cls
=
\frac{P(s + b)}{P(b)\hphantom{s +\;}}
=
\frac{
\prob{t \geq \tilde{t}}{s + b}
}{
\prob{t \geq \tilde{t}}{b}\hphantom{s +\;}
}
,
\end{equation}
in which the individual \pvalues\ are named
$\clb = P(b)$ and
$\clsb = P(s + b)$.%
\footnote{%
Despite the $\mathrm{CL}$ labels, these ($\cls$, $\clsb$, and $\clb$) are
not confidence levels ---
a confidence level $\gamma$ is a decision boundary used, for example,
to assign limits by solving for $x$ such that
$\gamma = 1 - P(x)$~\cite{pdg1996}.
In particle physics, \pvalues\ have sometimes been labelled $\mathrm{CL}$ since
at least 1984~\cite{pdg1984}, although textbooks from the '70s (which are cited
by~\cite{pdg1984}) do make the distinction~\cite{
eadie1971statistical,
frodesen1979probability
}.%
}%
\footnote{%
Some literature defines $\clb = 1 - p_b$~\cite{
Cowan:2010js,
pdg2022ynf
};
$p_b$ might not be a \pvalue\ if it excludes equality with the observed
test statistic.
}%
As discussed in Section~\ref{sec:searches_fisher}, \pvalues\ are likelihoods
on obfuscated data, so (with some caution) $\cls$ can be interpreted as a
likelihood ratio.%
\footnote{%
Likelihood ratios are useful if both their numerator and denominator use the
same data, so $\clsb$ and $\clb$ should use the same test statistic.
Identical test statistics are used for all results in presented in this thesis,
but differing test statistics have been allowed elsewhere~\cite{bock1998lower}.
}%
At a given confidence level $\gamma$ (typically $95\%$), limits are set to
values of $s$ for which $\cls = 1 - \gamma$.

\begin{figure}[tp]
\centering
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/searches_cls_plots_with_pvals_n.pdf}
\end{subfigure}
\caption[
A Poisson modelling example in the natural order.
]{%
A Poisson modelling example with $t(n) = \pm n$
test statistics, $b = 4.3$ and $s = 3.0$.
(left) \pvalues,
and (right) $\log$ likelihood ratios and $\log$ $\cls$ for
background and signal-plus-background models.
The $n$ test statistic is Neyman-Pearson optimal because it is monotonically
(linearly) related to the likelihood ratio.
Each $\cls$ is a \pvalue ratio.
By reducing the observed data to an inequality, the $p$-value selectively
censors some of its information.
}
\label{fig:searches_sb_n}
\end{figure}

Early forms of $\cls$ derived it from an identity with posterior tail
probabilities in Poisson or Gaussian special cases~\cite{
Helene1983upper,
pdg1988,
read2000modified,
pdg2022ynf
}
with a contrived frequentist interpretation~\cite{zech1988cls}.
For use in general cases, it is more recently described as a
``modified frequentist'',
``approximate confidence in the signal-only hypothesis''~\cite{
read2000modified,
Read2002cls
}.
High-profile usage in Higgs boson searches at the
Large Electron–Positron Collider~\cite{
read1997optimal,
bock1998lower,
etde1998prospects,
junk1999confidence,
lep2000searches,
lep2003search
}
appears to have propelled $\cls$ to prominence.
These early applications used data deficits to exclude their
background-plus signal models with the test statistic $t(n) = -n$, but
disliked how deficits led to small \pvalues\ for the background-only model.
Such \pvalues\ are illustrated in Figure~\ref{fig:searches_sb_n} (left);
in contrast, $\cls$ (shown on the right of this figure) can only favour the
background-only model.

\TODO{also show actual likelihood ratio}
\TODO{this implements a form of propaganda that speaks only truth while
censoring all information that supports alternative ideas}
\TODO{The one-sided nature of this evidence is paradoxical, but has
resolutions.}
\TODO{If an analysis were able to `is $n \geq 5$?', it could be answered in the
affirmative or negative, and would avoid this bias.}
\TODO{Otherwise, one can fall back to not trusting the reporter ---
the statements `\atlas\ reports $n \geq 0$' and `$n \geq 0$' are not identical
since the first might suggest that in fact $n = 0$, so wise likelihoods can be
could recover the original data.
One knows what the message says, but not whether it states the whole truth.
}

\begin{figure}[tp]
\centering
\begin{subfigure}{0.5\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/searches_cls_plots_with_pvals_order.pdf}
\caption{%
Ordering the data by a maximum likelihood ratio test statistic with $b=4.3$.
}
\end{subfigure}
\\[.5em]
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/searches_cls_plots_with_pvals_t.pdf}
\caption{%
(left) \pvalues\ for Poisson data,
and (right) $\log$ likelihood ratios and $\log$ $\cls$ for
background and signal-plus-background models.%
}
\end{subfigure}
\caption[
Analysis of a minimal Poisson modelling example with maximum likelihood ratio
test statistics
]{%
Analysis of a minimal Poisson modelling example with maximum likelihood ratio
test statistics $t(n) = \pm \Lambda(n)$,
where $\Lambda(n) = L(b)/\mathrm{max}_x\,L(x)$; $b = 4.3$ and $s = 3.0$.
This test statistic jumbles up the data into an that makes $n=11$ appear to be
weaker support for the signal than $n=9$.
Each $\cls$ is \pvalue\ ratio.
By reducing the observed data to an inequality, the $p$-value selectively
censors some of its information.
}
\label{fig:searches_sb_t}
\end{figure}

Figure~\ref{fig:searches_sb_t}

\TODO{refresh}
For example, maximum likelihood ratios (and variations on their theme) are
useful and commonly applied, but they can give \pvalues\ that appear
pathological --- consider a Poisson background ($b$) added to an
unconstrained signal yield ($s$).
A sensible interpretation would find that `excess' data well above the
background expectation favour the background-plus-signal hypothesis ($s + b$)
---
but if the signal is allowed to be negative, then a `deficit' of data below the
background can have a maximum likelihood ratio that favours $s + b$.
This scrambles excesses and deficits together in the same tail, so both get
small \pvalues, and so a small \pvalue\ implies neither a preference for the
signal hypothesis nor incompatibility with the background.
